Light inextensible strings (thread) under tension in the Schwarzschild geometry

### 2021-02-25 tpe_edreview_rebuttal.txt

> are measured in the same units.  Compare to Taylor & Wheeler's
> Exploring Black Holes, p. 2-28, Eq. (27).
>

Factor of M now added (actually this caused me some considerable
confusion and uncertainty.  The equation presented in Taylor and
Wheeler has a factor of 2M but it should be M---see the attached
mathematica screenshot, plus the identity arctanh(x) =
log((1+x)/(1-x))/2.

This was just a typo in that particular line and has no effect on any
other part of the manuscript.

> 7.  Suggestion: It appears that the $L$ and the $l$ that appear
> between Eq. (1) and Eq. (3) denote the same quantity.  Why use
> different symbols?
>

### 2021-02-25 string_physics_educator_revision3.tex
fixes #39
fixes #41
fixes #43

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
  \left.
  \sqrt{r(r-2M)} +\frac{1}{2}\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}.
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2M$ with no difficulty, even though the radial component of
the Schwarzschild metric approaches infinity there.  Elementary
methods show that the proper length $\ell$ of radial strings has some
non-Euclidean features.  Firstly, consider the proper length from
$r=2M$ to $r=2M(1+w)$ where $w>0, \left|w\right|\ll 1$.  Then

@@

\noindent where dashes denote differentiation with respect to~$\phi$.
The Euler-Lagrange equations for this system,
$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
  F}{\partial r}=0$ give an expression for the second derivative
of~$r$ (Mathematica idiom would be {\tt EulerEquations[Sqrt[y[x]\^{}2
      + y'[x]\^{}2/(1-1/y[x])],y[x],x]}):

\begin{equation}\label{rdashdash}
  r''\left(\phi\right) =
  (r-2M) + \frac{(2r-3M)\left(r'\right)^2}{r\left(r-2M\right)},\qquad r>2M.
\end{equation}

\noindent Solutions to equation~\ref{rdashdash} correspond to a taut
string confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rdashdash} with the corresponding equation for
equatorial null geodesics, which is usually given~\cite{wald} in terms
of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that
light, being sensitive to the time component of the metric, behaves
differently from string, which is not (at this point it might be worth
``reminding'' students that a straight line in flat space,
$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
$r''=r+2\left(r'\right)^2/r$).

@@

There does not appear to be a simple analytical solution to
equation~\ref{rdashdash}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $r'$ is small, then a string
follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
Working to first order, and writing~$d$ for the difference in
Schwarzschild~$r$ between string and null geodesic, we have $d''=M$.
Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with boundary
conditions corresponding to both string and null geodesic being
tangential at the $(x,y)$ origin, we see that
$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
the string would measure
$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
equation~\ref{schwarzschild}---and obtain
$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
which would correspond to a radius of curvature
of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
taut string would {\em look} curved to a static observer in
Schwarzschild coordinates, in the sense that a tangential light beam
and taut string would diverge quadratically.  At the photon sphere
$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
For a solar mass black hole this would be about $15\unit{km}$, a
readily understandable result.

@@

Here I solve equation~\ref{rdashdash} numerically, using the R
programming language~\cite{rcore2019} and the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  First we
consider strings for which $r>2M$ everywhere (that is, completely
exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rdashdash}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.

>>>>

\begin{equation}\label{radial_string_length}
  \int_{r=r_1}^{r_2}\frac{dr}{\sqrt{1-2M/r}}=
  \left.
  \sqrt{r(r-2M)} +M\log\left(
  \frac{\sqrt{r}+\sqrt{r-2M}}{\sqrt{r}-\sqrt{r-2M}}\right)
  \right|_{r_1}^{r_2}.
  \end{equation}

Observe that integration may be performed with a lower limit
of~$r_1=2M$ with no difficulty, even though the radial component of
the Schwarzschild metric approaches infinity there.  Elementary
methods show that the proper length $\ell$ of radial strings has some
non-Euclidean features.  Firstly, consider the proper length from
$r=2M$ to $r=2M(1+w)$ where $w$ is small, $0<w\ll 1$.  Then

@@

\noindent where primes denote differentiation with respect to~$\phi$.
The Euler-Lagrange equations for this system,
$\frac{d}{d\phi}\frac{\partial F}{\partial r'}-\frac{\partial
  F}{\partial r}=0$ give an expression for the second derivative
of~$r$ (Mathematica idiom would be {\tt EulerEquations[Sqrt[y[x]\^{}2
      + y'[x]\^{}2/(1-1/y[x])],y[x],x]}):

\begin{equation}\label{rprimeprime}
  r''\left(\phi\right) =
  (r-2M) + \frac{(2r-3M)\left(r'\right)^2}{r\left(r-2M\right)},\qquad r>2M.
\end{equation}

\noindent Solutions to equation~\ref{rprimeprime} correspond to a taut
string confined to Flamm's paraboloid~\cite{flamm1916}; see
Figure~\ref{flamm}.  It is interesting to compare
equation~\ref{rprimeprime} with the corresponding equation for
equatorial null geodesics, which is usually given~\cite{wald} in terms
of $u=2M/r$ as $u''=3u^2/2-u$, but is equivalent
to~$r''\left(\phi\right)=r-3M+2\left(r'\right)^2/r$.  This shows that
light, being sensitive to the time component of the metric, behaves
differently from string, which is not (at this point it might be worth
``reminding'' students that a straight line in flat space,
$r=r_0\sec\left(\phi-\phi_0\right)$, obeys
$r''=r+2\left(r'\right)^2/r$).

@@

There does not appear to be a simple analytical solution to
equation~\ref{rprimeprime}; but it is a straightforward second order
ODE, amenable to numerical solution and no harder than the equivalent
for null geodesics, a standard homework problem.  Here we consider
only strings exterior to the event horizon, $r>2M$, which have a
well-defined point of closest approach to the singularity.  Because
the system is invariant under~$\phi\longrightarrow-\phi$, the string
must be symmetrical about this point.  If $r'$ is small, then a string
follows $r''=r-2M$, while tangential null geodesics follow $r''=r-3M$.
Working to first order, and writing~$d$ for the difference in
Schwarzschild~$r$ between string and null geodesic, we have $d''=M$.
Transforming coordinates to $(x,y) = (r\phi,r-r_0)$ and with boundary
conditions corresponding to both string and null geodesic being
tangential at the $(x,y)$ origin, we see that
$d=\frac{M}{2}(x/r_0)^2$.  However, a static local observer tied to
the string would measure
$d_\mathrm{local}=d/\sqrt{1-2M/r_0}$---applying
equation~\ref{schwarzschild}---and obtain
$d_\mathrm{local}=\frac{M}{2}(x/r_0)^2\left(1-2M/r_0\right)^{-1/2}$
which would correspond to a radius of curvature
of~$r_0\cdot\frac{r_0}{M}\cdot\left(1-2M/r_0\right)^{-1/2}$.  Thus a
taut string would {\em look} curved to a static observer in
Schwarzschild coordinates, in the sense that a tangential light beam
and taut string would diverge quadratically.  At the photon sphere
$r_0=3M$, such a string would appear to have a radius of~$\sqrt{27}M$.
For a solar mass black hole this would be about $15\unit{km}$, a
readily understandable result.

@@

Here I solve equation~\ref{rprimeprime} numerically, using the R
programming language~\cite{rcore2019} and the \verb+deSolve+ package
for ordinary differential equations~\cite{soetart2010}.  First we
consider strings for which $r>2M$ everywhere (that is, completely
exterior strings).
Figure~\ref{closest_approach_non_self_intersecting} shows a sequence
of non-self-intersecting numerical solutions to
equation~\ref{rprimeprime}, rotated so that the tangential point of
closest approach occurs at~$\phi=0$.
Figure~\ref{closest_approach_self_intersecting} shows the
corresponding diagram for self-interesecting strings
(Figure~\ref{light_closest_approach} shows equivalent null geodesics
for comparison).  A bisection technique reveals that if the closest
approach is greater than a critical value of about~$\simeq 1.0761\cdot
2M$, the string does not self-intersect; if the closest approach is
less than this value, then the string crosses itself at~$\phi=\pi/2$.
